The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2018-12-22 Paper.Rmd
further edits

Question Q1 was answered principally via analysis of the
data using a model selection approach. For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, as in the Introduction, we produced ranks 
$(\tilde{u}_i,\tilde{v}_i)$ where $\tilde{u}_i$ is the rank of $x_i$ in the set 
$\{x_i:i=1,...,n\}$ and $\tilde{v}_i$ is the rank of $y_i$ in the set $\{y_i:i=1,...,n\}$ and 
we set $u_i=\tilde{u}_i/(n+1)$ and $v_i=\tilde{v}_i/(n+1)$; subsequent analysis used these 
*normalized ranks* $u_i$ and $v_i$. Ranks were $1$ for the smallest element in a set. 
Second, we did a test of the independence of the $u_i$ and $v_i$ (implemented in `BiCopIndTest` in the `VineCopula` package). 
<!--***DAN: X and Y to be flled in, can't remember right now what they are and don't want to stop
to look it up, Shya, you could also fill these in if you get a chance
***Shya: Done.-->
Subsequent model 
selection algorithms are ineffective if data could not be distinguished from independent data 
since most or all bivariate copula families considered include the independent copula in the 
family or as a boundary case. If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the bivariate copula families listed previously to the ranked data 
$(u_i,v_i)$ via maximum likelihood, using the `BiCopEst` function in `VineCopula`, and 
obtained AIC and BIC values and accompanying model weights, $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. The weight of a model represents the 
relative support of data for the model; weights sum to 1 across models considered. The lower- and 
upper-tail dependence of the best-fitting member of each family was also provided by `BiCopEst`. 
Fourth, $\text{AIC}_{\text{w}}$ values for all fitted models were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods will give the *relative* support of several models, but will not,
on their own, indicate whether any of the models selected is an objectively good fit: all 
models considered might poorly describe the data. To test this, and as the final step of
our procedure, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure implemented in `BiCopGofTest` of `VineCopula`. The method performed 
two tests, one based on a Cramer-von Mises statistic, and one based on a 
Kolmogorov-Smirnov statistic. For keep computation times reasonable, 
an initial run using $100$ bootstraps was performed, and if
the $p$-value resulting from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

@@

We also defined the statistic $\Ps_{l_b,u_b}$ (Appendix \ref{SM-nonparam_stats}), 
which has a similar 
interpretation to $\cor_{l_{b_k},u_{b_k}}(u,v)$; we use both statistics 
because in some instances $\Ps_{l_b,u_b}$ appears to have more power to
reveal tail dependence than does $\cor_{l_{b_k},u_{b_k}}(u,v)$. 
<!--***DAN: need to make this a bit more precise, perhaps just enough to cite one of the supp mat sections where tests are done of these stats, but you need to review the tests first and what they show-->
Our third statistic, $\Dsq_{l_b,u_b}(u,v)$, is the average squared 
distance between points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ and the line $v=u$.
<!--***DAN: Shyamolina, is this actually right, i.e., did we use the average, or did we 
use the sum divided by n? Whichever way we did it is OK, but please check the code and revise the above 
text if necessary and insert a comment answering this comment. I tried to check this, myself, 
but I was having a hard time finding where the statistics are implemented. Please also
indicate in your response comment what line of what file has the implementation of these
statistics.
***Shya: In code we used the sum of squared distances divided by n(=number of points lying within two boundaries) which I think okay to be called as "average-squared distance" or (to be precised) average of squared distances of points from 1:1 line.
Reference : CopulaFunctions_flexible.R file, line 24.
***DAN: OK, good, I am leaving these comments here as a reminder-->
Unlike $\cor_{l_b,u_b}$ and 
$\Ps_{l_b,u_b}$, which are large when points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ cluster close to the line $v=u$, $\Dsq_{l_b,u_b}$ is 
small in that case. 
Thus large values of $\cor_{l_b,u_b}$ and $\Ps_{l_b,u_b}$ indicate strong 
dependence in the portions of the distributions of $u$ and $v$ given by 
the bounds, whereas small values of 
$\Dsq_{l_b,u_b}$ indicate strong dependence.

@@

<!--bivariate example-->
We discuss the example of the soil C and N data (see Data), pictured in 
Fig. \ref{fig_biv_multi_raw_cop}A (raw data) and E (copula plot). The variables were 
non-independent ($p=$ `r BivMS_res_Loecke$IndepTestRes`, to within the precision available 
from `BiCopIndTest` in the `VineCopula` package); non-independence is also visually apparent
(Fig. \ref{fig_biv_multi_raw_cop}E).
The Kendall correlation was `r round(BivMS_res_Loecke$TauVal,2)`. We fitted the 16 copulas
mentioned in the previous section to the 
data pictured in Fig. \ref{fig_biv_multi_raw_cop}E; AIC and BIC weights and corresponding
lower- and upper-tail statistics for the fitted distributions were computed (Table \ref{SM-tab_soilCNfit}). 
`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]` was the 
minimum-AIC copula, with AIC weight `r round(max(BivMS_res_Loecke$InfCritRes$AICw),2)`, whereas the 
normal copula had AIC weight essentially 0, answering the 
first part of Q1 for these data. 
The model-averaged lower- and upper-tail dependence statistics were `r round(BivMS_res_Loecke$relLTdep_AICw,3)` 
and `r round(BivMS_res_Loecke$relUTdep_AICw,3)`, respectively (AIC weights were used for model averaging). 
These values are distinct from what a normal copula would give, namely, 0, answering the second part 
of Q1 for these data. The values are also substantially 
distinct from each other, answering the third part of Q1.

<!--***DAN: made it to here 2018 12 20 920pm-->

Although these results do convincingly show non-normal copula structure, 
we caution that the model-averaged tail-dependence statistics may be biased 
because even the lowest-AIC copula 
(`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]`) 
was a poor fit ($p=$ `r BivMS_res_Loecke$GofRes_CvM` and `r BivMS_res_Loecke$GofRes_KS` for the 
Cramer-von Mises and Kolmogorov-Smirnov goodness-of-fit tests, respectively, to within the precision 
available from `BiCopGofTest` in the `VineCopula` package). However, nonparametric statistics verified
both the asymmetry of tail dependence for the soil data and the conclusion that tail dependence 
in the data was distinct from a normal copula (Fig. \ref{SM-fig_soilCN_nonparam}, Table \ref{SM-tab_soilCN_asym}).

<!--***DAN: the above block of text is the first paart of a para, and i have edited it. The below
block is the rest of the same para, and I have not-->

where each of our statistics ($\cor$, $\Ps$, $\Dsq$), as computed for the ranges $0-0.1$, $0.1-0.2$, ..., $0.9-1$ (with $10$ number of bins), was compared to the distributions of values of the same statistic with the same ranges computed on 1000 Normal-copula surrogate datasets with similar Kendall or Spearman correlations, in different runs for Kendall and Spearman (Fig. \ref{SM-fig_soilCN_nonparam}). Each bin in each subplot of Fig. \ref{SM-fig_soilCN_nonparam} contains a number indicating how many Normal surrogates out of 1000, are different ($95\%$ confidence interval) from Normal-copula statistics. Those numbers clearly show that tail dependence of data is stronger in both the lower and upper tails than would be expected under a null hypothesis of a normal copula, answering the second part of \hyperref[Intro]{\textbf{(Q1)}}.

Furthermore, the difference of model-averaged tail dependence statistics (lower minus upper), `r BivMS_res_Loecke$relLTdep_AICw-BivMS_res_Loecke$relUTdep_AICw`, was different from 0, answering the third part of \hyperref[Intro]{\textbf{(Q1)}}. To be precise, this difference being negative, indicated stronger upper-tail dependence than lower, which was also supported by non-parametric results (see Table \ref{SM-tab_soilCN_asym}).

\begin{figure}
\begin{center}
\includegraphics[width=6.5in]{./Results/BivarDataPlot.pdf}
\caption{Upper panels shows raw data plots for (A) soil C and N data, (B) $log_{10}$(basal metabolic rate (BMR)) vs. log(body mass) 
for birds, (C) log(BMR) vs. log(body mass) for mammals, and (D) above-ground plant biomass vs. Shannon's index (H) from Cedar Creek. Bottom panels show the corresponding copula plots. \label{fig_biv_multi_raw_cop}}
\end{center}
\end{figure}
<!--***DAN: Shya, we need to produce a single-pdf figure using R, not use latex to
cobble together separate pdfs for each panel. Also, I think we can dispense with the 
plots of the multivariate data. I have created a script that will do both of these things, 
and inserted a call to it in the sup mat Rmd. Right now it only plots dummy, filler data. 
Please look at MakeBivarDataPlot.R (that's the script) and search for
"Shya", and do as it says in those comments, so the actual data are imported into that
script and plotted. I have changed the above so this new figure is plotted instead of the
cobbled together old figures.
***Shya: filled in MakeBivDataPlot.R : Done.-->

@@

dt <- tibble(Description = c("1. P-value, independence test",
           "2. Minimum-AIC copula",
           "3. AIC weight of minimum-AIC copula",
           "4. AIC weight of N copula",
           "5. P-value, Cramer-von Mises goodness of fit test",
           "6. P-value, Kolmogorov-Smirnov goodness of fit test",
           "7. Model averaged lower-tail dependence",
           "8. Model averaged upper-tail dependence",
           "9. Model averaged lower- minus upper-tail dependence",
           "10. Asymmetry rank in surrogates")
)
tab_bivar_summary<-cbind(dt,tab_bivar_summary)
knitr::kable(tab_bivar_summary, 
             format="latex",
             caption = "Summary of Q1 results for bivariate datasets. The $p$-values in rows 5-6 and are for the minimum-AIC copula. Model averaging used for rows 7-9 was based on AIC weights. For row 10, the statistic $P_l-P_u$ was computed for real data and for 1000 surrogate datasets with N copulas, and the rank of real-data value in the distribution of surrogate values is presented. \\label{tab_bivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ", "Soil C and N", "Bird masses and BMR", "Mammal masses and BMR", "Cedar creek data")
             )%>%column_spec(2:5,width="1.4cm")%>%
             column_spec(1, width = "8.5cm")%>%kable_styling(latex_options = "basic",position = "center")
```
<!--***DAN: Shya, it says Pl-Pu above, but I suspect it was P_first slice - P_last slice, right?
Can you please look back at the code, if necessary, and write a comment below this one 
indicating what results are actually presented in the table?
***Shya: Yes, you are right, here Pl-Pu means P_first slice - P_last slice. In table, item 10 : Assymetry rank is shown for Kendall preserving surroagates.
Reference : bivfunctionplot.R file, lines 134,135,161-->

>>>>

Question Q1 was answered principally via analysis of the
data using a model selection approach. For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, as in the Introduction, we produced ranks 
$(\tilde{u}_i,\tilde{v}_i)$ where $\tilde{u}_i$ is the rank of $x_i$ in the set 
$\{x_i:i=1,...,n\}$ and $\tilde{v}_i$ is the rank of $y_i$ in the set $\{y_i:i=1,...,n\}$ and 
we set $u_i=\tilde{u}_i/(n+1)$ and $v_i=\tilde{v}_i/(n+1)$; subsequent analysis used these 
*normalized ranks* $u_i$ and $v_i$. Ranks were $1$ for the smallest element in a set. 
Second, we did a test of the independence of the $u_i$ and $v_i$ 
(implemented in `BiCopIndTest` in the `VineCopula` package). 
Subsequent model 
selection algorithms are ineffective if data could not be distinguished from independent data 
since most or all bivariate copula families considered include the independent copula in the 
family or as a boundary case. If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the bivariate copula families listed previously to the ranked data 
$(u_i,v_i)$ via maximum likelihood, using the `BiCopEst` function in `VineCopula`, and 
obtained AIC and BIC values and accompanying model weights, $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. The weight of a model represents the 
relative support of data for the model; weights sum to 1 across models considered. The lower- and 
upper-tail dependence of the best-fitting member of each family was also provided by `BiCopEst`. 
Fourth, $\text{AIC}_{\text{w}}$ values for all fitted models were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods will give the *relative* support of several models, but will not,
on their own, indicate whether any of the models selected is an objectively good fit: all 
models considered might poorly describe the data. To test this, and as the final step of
our procedure, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure implemented in `BiCopGofTest` of `VineCopula`. The method performed 
two tests, one based on a Cramer-von Mises statistic, and one based on a 
Kolmogorov-Smirnov statistic. For keep computation times reasonable, 
an initial run using $100$ bootstraps was performed, and if
the $p$-value resulting from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

@@

We also defined the statistic $\Ps_{l_b,u_b}$ (Appendix \ref{SM-nonparam_stats}), 
which has a similar 
interpretation to $\cor_{l_{b_k},u_{b_k}}(u,v)$; we use both statistics 
because $\cor_{l_{b_k},u_{b_k}}(u,v)$ is more familiar, but 
in some instances $\Ps_{l_b,u_b}$ appears to have more power to
reveal tail dependence than does $\cor_{l_{b_k},u_{b_k}}(u,v)$. 
<!--***DAN: need to make this a bit more precise, perhaps just enough to cite one of the supp mat sections where tests are done of these stats, but you need to review the tests first and what they show-->
Our third statistic, $\Dsq_{l_b,u_b}(u,v)$, is the average squared 
distance between points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ and the line $v=u$.
<!--***DAN: Shyamolina, is this actually right, i.e., did we use the average, or did we 
use the sum divided by n? Whichever way we did it is OK, but please check the code and revise the above 
text if necessary and insert a comment answering this comment. I tried to check this, myself, 
but I was having a hard time finding where the statistics are implemented. Please also
indicate in your response comment what line of what file has the implementation of these
statistics.
***Shya: In code we used the sum of squared distances divided by n(=number of points lying within two boundaries) which I think okay to be called as "average-squared distance" or (to be precised) average of squared distances of points from 1:1 line.
Reference : CopulaFunctions_flexible.R file, line 24.
***DAN: OK, good, I am leaving these comments here as a reminder-->
Unlike $\cor_{l_b,u_b}$ and 
$\Ps_{l_b,u_b}$, which are large when points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ cluster close to the line $v=u$, $\Dsq_{l_b,u_b}$ is 
small in that case. 
Thus large values of $\cor_{l_b,u_b}$ and $\Ps_{l_b,u_b}$ indicate strong 
dependence in the portions of the distributions of $u$ and $v$ given by 
the bounds, whereas small values of 
$\Dsq_{l_b,u_b}$ indicate strong dependence.

@@

<!--bivariate example-->
We discuss the example of the soil C and N data (see Data), pictured in 
Fig. \ref{fig_biv_multi_raw_cop}A (raw data) and E (copula plot). The variables were 
non-independent ($p=$ `r BivMS_res_Loecke$IndepTestRes`, to within the precision available 
from `BiCopIndTest` in the `VineCopula` package); non-independence is also visually apparent
(Fig. \ref{fig_biv_multi_raw_cop}A,E).
The Kendall correlation was `r round(BivMS_res_Loecke$TauVal,2)`. We fitted the 16 copulas
mentioned in the previous section to the 
data pictured in Fig. \ref{fig_biv_multi_raw_cop}E; AIC and BIC weights and corresponding
lower- and upper-tail statistics for the fitted distributions were computed (Table \ref{SM-tab_soilCNfit}). 
`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]` was the 
minimum-AIC copula, with AIC weight `r round(max(BivMS_res_Loecke$InfCritRes$AICw),2)`, whereas the 
normal copula had AIC weight essentially 0, answering the 
first part of Q1 for these data. 
The model-averaged lower- and upper-tail dependence statistics were `r round(BivMS_res_Loecke$relLTdep_AICw,3)` 
and `r round(BivMS_res_Loecke$relUTdep_AICw,3)`, respectively (AIC weights were used for model averaging). 
These values are distinct from what a normal copula would give, namely, 0, answering the second part 
of Q1 for these data. The values also differed substantially 
from each other, answering the third part of Q1. The numbers reflect stronger upper- than lower-tail
dependence for the soil C and N data, and this can be seen visually in the extreme tails 
of the copula plot (Fig. \ref{fig_biv_multi_raw_cop}E).

\begin{figure}
\begin{center}
\includegraphics[width=6.5in]{./Results/BivarDataPlot.pdf}
\caption{Upper panels show raw data plots for (A) soil C and N data, (B) $\log_{10}$(basal metabolic rate (BMR)) vs. $\log_{10}$(body mass) 
for birds, (C) $\log_{10}$(BMR) vs. $\log_{10}$(body mass) for mammals, and (D) above-ground plant biomass vs. Shannon's index (H) from Cedar Creek. Bottom panels show the corresponding copula plots. \label{fig_biv_multi_raw_cop}}
\end{center}
\end{figure}

Although these results do convincingly show non-normal copula structure, 
we caution that the model-averaged tail-dependence statistics may be biased 
because even the lowest-AIC copula 
(`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]`) 
was a poor fit ($p=$ `r BivMS_res_Loecke$GofRes_CvM` and `r BivMS_res_Loecke$GofRes_KS` for the 
Cramer-von Mises and Kolmogorov-Smirnov goodness-of-fit tests, respectively, to within the precision 
available from `BiCopGofTest` in the `VineCopula` package). However, nonparametric statistics verified
both that tail dependence was distinct from that of a normal copula for the soil data, and that it 
was asymmetric (Fig. \ref{SM-fig_soilCN_nonparam}, Table \ref{SM-tab_soilCN_asym}). 
The values of the statistics $\cor_{l_b,u_b}$, $\Ps_{l_b,u_b}$ and $\Dsq_{l_b,u_b}$ for real data
were compared to distributions of their values on 1000 Kendall- or Spearman-preserving N surrogates
of the data,
separately in two comparisons for each of the ranges $(l_b,u_b)=(0,0.1),\ldots,(0.9,1)$ 
(Fig. \ref{SM-fig_soilCN_nonparam}). Results show that tail dependence of data is stronger in 
both the lower and upper tails than would be expected under a null hypothesis of a normal copula.
The values of the statistics $\cor_{0,0.1}-\cor_{0.9,1}$, $\Ps_{0,0.1}-\Ps_{0.9,1}$ and
$\Dsq_{0.9,1}-\Dsq_{0,0.1}$ for real data were also compared to distributions of their values
on 1000 Kendall- or Spearman-preserving N surrogates, separately in two comparisons for each
of the $(l_b,u_b)$ ranges listed above (Table \ref{SM-tab_soilCN_asym}). Results showed that
upper-tail dependence was significantly stronger than lower-tail dependence.

<!--***DAN: made it to here 2018 12 21 at 10pm-->

@@

dt <- tibble(Description = c("1. P-value, independence test",
           "2. Minimum-AIC copula",
           "3. AIC weight of minimum-AIC copula",
           "4. AIC weight of N copula",
           "5. P-value, Cramer-von Mises goodness of fit test",
           "6. P-value, Kolmogorov-Smirnov goodness of fit test",
           "7. Model averaged lower-tail dependence",
           "8. Model averaged upper-tail dependence",
           "9. Model averaged lower- minus upper-tail dependence",
           "10. Asymmetry rank in surrogates")
)
tab_bivar_summary<-cbind(dt,tab_bivar_summary)
knitr::kable(tab_bivar_summary, 
             format="latex",
             caption = "Summary of Q1 results for bivariate datasets. The $p$-values in rows 5-6 are for the minimum-AIC copula. Model averaging used for rows 7-9 was based on AIC weights. For row 10, the statistic $P_{0,u_b}-P_{1-u_b}$ was computed for real data and for 1000 Kendall-preserving surrogate datasets with N copulas, and the rank of real-data value in the distribution of surrogate values is presented. The value of $u_b$ depended on the size of the dataset, and can be found in Figs. \\ref{SM-fig_soilCN_nonparam} to \\ref{SM-fig_cc2000_nonparam} and Tables \\ref{SM-tab_soilCN_asym}, \\ref{SM-tab_bmr_birds_asym}, \\ref{SM-tab_bmr_mammals_asym}, and \\ref{SM-tab_cc_asym_2000}. \\label{tab_bivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ", "Soil C and N", "Bird masses and BMR", "Mammal masses and BMR", "Cedar Creek data")
             )%>%column_spec(2:5,width="1.4cm")%>%
             column_spec(1, width = "8.5cm")%>%kable_styling(latex_options = "basic",position = "center")
```
<!--***DAN: Shya, it says Pl-Pu above, but I suspect it was P_first slice - P_last slice, right?
Can you please look back at the code, if necessary, and write a comment below this one 
indicating what results are actually presented in the table?
***Shya: Yes, you are right, here Pl-Pu means P_first slice - P_last slice. In table, item 10 : Assymetry rank is shown for Kendall preserving surroagates.
Reference : bivfunctionplot.R file, lines 134,135,161
***DAN: OK, great. I have made changes to deal with this, but am leaving the comment for now
as a reminder that we looked at this-->
